---
title: "NMAI061-22-EX2"
author: 'Matej Nemec'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1)Poisson distribution MLE formula derivation
Probability density function for Poisson distribution:
\[P(X=x)=\frac{\lambda^xe^-\lambda}{x!} \]
Likelihood function:
\[L(\lambda;x_1,...,x_n)=\prod_{i=1}^{n}{\frac{\lambda^{x_i}e^-\lambda}{x_i!}}\]
We can use natural logarithm to simplify so we are able to replace product with sum, this gives log likelihood function:
\[l(\lambda;x_1,...,x_n)=ln(\prod_{i=1}^{n}{\frac{\lambda^{x_i}e^-\lambda}{x_i!}})\]
Which can be written as:
\[l(\lambda;x_1,...,x_n)=-n\lambda+ln(\lambda)\sum_{i=1}^{n}{{x_i}}-\sum_{i=1}^{n}{ln({x_i!}})\]
Next we calculate the derivative of the log likelihood function w.r.t. $\lambda$ and put it equal to 0:
\[\frac{d}{d\lambda}l(\lambda;x_1,...,x_n)=-n+\frac{1}{\lambda}\sum_{i=1}^{n}=0\]
Finally we solve for $\lambda$ leaving us with MLE for $\lambda$ equal to sample mean:
\[\lambda=\frac{1}{n}\sum_{i=1}^{n}{x_i}\]

## 2)MLE for simulated data
We keep the $\lambda$ of 9 from previous excercise and randomly sample from the corresponding Poisson distribution. We sample only 20 times to make it somewhat interesting:
```{r}
n=20
l=9
sample=rpois(n=n,lambda=l)
cat(paste0('Sample mean for ',n,' random samples with lambda ',l,' is ',mean(sample),'.'))
```

## 3)Reapeated sampling and estimate distribution

Histogram for $\lambda$ estimates from repeated random sampling:

```{r}
repeats=1000
simulation=matrix(rpois(n*repeats,lambda=l),nrow=repeats)
estimates=rowSums(simulation)/n
par(mfrow=c(1,2))
hist(estimates,prob=TRUE,breaks = seq(min(estimates)-sd(estimates), 
max(estimates)+sd(estimates),by = sd(estimates)/100),xlab='')
log_estimates=log(estimates)
hist(log_estimates,prob=TRUE,breaks = seq(min(log_estimates)-sd(log_estimates),
max(log_estimates)+sd(log_estimates), by = sd(log_estimates)/100),xlab='',ylab = '')
```
\newpage

## 4)Confidence interval estimation and testing

We can see that $\lambda$ of 9 falls within confidence intervals for all $\lambda$ estimations.
(Note: We only visualize 200 randomly picked estimations from 1000 for the graph to be readable):

```{r}
margin=qnorm(0.975)*estimates/sqrt(20)
lower_cis=estimates-margin
upper_cis=estimates+margin
plot(NULL,ylim=c(min(estimates)-8*sd(estimates),max(estimates)+8*sd(estimates)),
xlim=c(0,200),ylab='',xlab='',main='CIs for lambda')
i=1
for(k in sample(1:1000, 200)) {
    lines(x=c(i,i),y=c(lower_cis[k],upper_cis[k]))
    i=i+1
}
abline(h=9)
```